The TFR-HDI controversy: an example Bayesian data analysis with PyMC

### 2011-07-05 tex/paper.tex
add nonlinear model to writeup

\title{An example Bayesian data analysis with PyMC:\\The TFR-HDI controversy}

@@

        \paragraph*{Methods:} I replicated several analyses of the relationship between Human Development Index (HDI) and Total Fertility Rate (TFR) using the PyMC package for Python.

@@

\section*{Background}
Bayesian data analysis is an important tool, but it has a steep
learning curve.  MCMC packages like BUGS and PyMC can reduce the
barrier to entry significantly, by removing the burden of coding an
MCMC sampler (e.g. Gibbs sampler) from scratch.  However these systems
still have overhead of their own.

This paper and accompanying source code provide a complete (small)
example of a Bayesian data analysis with PyMC.  In addition to showing
an example of the PyMC syntax necessary, I have also described the
elements of my typical workflow in building, fitting, and validating a
model, as well as some elements of test-driven software development
and debugging PyMC code with the Python debugger.

@@

\section*{Methods}
  \subsection*{PyMC project structure}
  My template for a Bayesian data analysis with PyMC separates the
  source code into 4 Python files, which do not correspond precisely
  to the four stages of Bayesian data analysis: data preparation,
  model building, model fitting, model checking.

@@

  \subsection*{Model Building}
I started with the simplest relationship between HDI and TFR, which is
 a "linear model" that posits for a given increment in HDI there is a
 corresponding decrement in TFR, and formalizes this according to the
 equations
\begin{align*}
\TFR_i &= \beta_0 + \beta_1 \HDI_i + \epsilon_i\\
\epsilon_i &\sim N(0, \sigma^2)
\end{align*}

>>>>

\title{The TFR-HDI controversy: an example Bayesian data analysis with PyMC}

@@

        \paragraph*{Methods:} I replicated part of an analyses of the relationship between Human Development Index (HDI) and Total Fertility Rate (TFR) using the PyMC package for Python.

@@

\section*{Background}
Bayesian data analysis is an important tool, but it has a steep
learning curve.  MCMC packages like BUGS \ref{BUGS} and PyMC
\ref{PyMC} can reduce the barrier to entry significantly, by removing
the burden of coding an MCMC sampler (e.g. Gibbs sampler) from
scratch.  However these systems still have overhead of their own.

This paper and accompanying source code \ref{Code} provide a complete
(small) example of a Bayesian data analysis with PyMC.  In addition to
showing an example of the PyMC syntax necessary, I have also described
the elements of my typical workflow in building, fitting, and
validating a model, as well as some elements of test-driven software
development and debugging PyMC code with the Python debugger.

@@

\section*{Methods}
  \subsection*{PyMC project structure}
  My template for a Bayesian data analysis with PyMC
  \ref{PyMC-template} separates the source code into 4 Python files,
  which do not correspond precisely to the four stages of Bayesian
  data analysis: data preparation, model building, model fitting,
  model checking.

@@

  \subsection*{Model Building}
I started with the simplest relationship between HDI and TFR, which is
a ``linear model'' that posits for a given increment in HDI there is a
corresponding decrement in TFR, and formalizes this according to the
equations
\begin{align*}
\text{Linear Model (LM):}\qquad
\TFR_i &= \beta_0 + \beta_1 \HDI_i + \epsilon_i\\
\epsilon_i &\sim N(0, \sigma^2)
\end{align*}

Implementing this in PyMC may seem hardly worth the trouble, since
this sort of ordinary least squares model has a simple closed-form
solution, as well as numerous implementations, like the regress
command in STATA.  It will be clear why this is worthwhile after
modifying the systematic part to yield a (slightly) more complicated
nonlinear model.

\begin{align*}
\text{Nonlinear Model (NM):}\qquad
\TFR_i &= \beta_0 + \beta_1 \HDI_i + \beta_2(\HDI_i - \gamma)^+ \ + \epsilon_i,\\
\epsilon_i &\sim N(0, \sigma^2),
\end{align*}
where $(\HDI_i - \gamma)^+$ equals $\HDI_i - \gamma$ if this quantity
is positive, and $0$ otherwise.
